---
title: "Basic Diabetes Distribution"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(plotly)
library(maps)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

diabetes_df = read_csv("data/cleaned_diabetes_data.csv") # |> 
  # mutate(diab_type = factor(diab_type, levels = c(1, 2), labels = c("Type 1", "Type 2")))
```


# Defining Diabetes {.tabset}

**`has_diabetes`**: This variable describes the people in the survey who responded after being asked the question, "Have you ever been told you had diabetes?" 

**`diab_type`**: This variable comes directly from the dataset, where respondents who were diabetic per the `has_diabetes` question were asked "What type of diabetes do you have?"

**`eval_type`**: Due to the missing data for `diab_type`, we created another variable `eval_type` that classifies a respondent to have type II diabetes if the respondent was older than 30, not pregnant, and were diabetic as per the `has_diabetes` question. We classified a respondent to have type I diabetes if the respondent was younger than 30 and were diabetic as per the `has_diabetes` question.


```{r diabetes_distributions}
diabetes_dist = function(df, var) {
  df |> 
    group_by({{ var }}) |> 
    summarize(count = n()) |> 
    knitr::kable()
}

diabetes_plot = function(df, var) {
  diabetes_df |> 
    ggplot(aes(x = {{ var }})) +
    geom_bar(aes(fill = factor({{ var }})), color = "black", na.rm = FALSE) + 
    geom_text(
      stat = "count",  # Use the count statistic for frequencies
      aes(label = ..count..),  # Access the count directly
      vjust = -0.5,  # Adjust position of labels above the bars
      na.rm = FALSE
    ) + 
    labs(
      title = "Distribution of Diabetes", 
      x = "Diabetes Status", 
      y = "Count", 
      fill = as.character(rlang::ensym(var))
    ) +
    theme_minimal() +
    theme(legend.position="bottom", legend.title = element_blank())
}
```

## Diabetes Status (`has_diabetes`) {.tabset}

### Table

```{r message=FALSE, warning=FALSE}
diabetes_dist(df = diabetes_df, var = has_diabetes)
```

### Graph

```{r message=FALSE, warning=FALSE}
diabetes_plot(df = diabetes_df, var = has_diabetes)
```

## Reported Type (`diab_type`) {.tabset}

### Table

```{r message=FALSE, warning=FALSE}
diabetes_dist(df = diabetes_df, var = diab_type)
```

### Graph

```{r message=FALSE, warning=FALSE}
diabetes_plot(df = diabetes_df, var = diab_type)
```

## Evaluated Type (`eval_type`) {.tabset}

### Table

```{r message=FALSE, warning=FALSE}
diabetes_dist(df = diabetes_df, var = eval_type)
```

### Graph

```{r message=FALSE, warning=FALSE}
diabetes_plot(df = diabetes_df, var = eval_type)
```

# Diabetes in the U.S.

```{r us_dist, message = FALSE, warning = FALSE}

# Ensure state names/abbreviations match plotly's expected format
state_abbreviation_map =
  tibble(
    state = c(tolower(state.name), "district of columbia", "puerto rico", "guam", "virgin islands"),
    abb = c(state.abb, "DC", "PR", "GU", "VI")
  )

diabetes_by_state = 
  diabetes_df |> 
  mutate(state = tolower(state)) |>  # Convert state names to lowercase
  group_by(state) |> 
  summarize(
    prevalence = round((sum(has_diabetes == "Diabetic", na.rm = TRUE) / n()) * 100, 2), # Calculate prevalence
    .groups = "drop"
  ) |> 
  complete(
    state = state_abbreviation_map$state,
    fill = list(prevalence = NA)  # Fill missing states/territories with NA
  ) |> 
  left_join(state_abbreviation_map, by = "state")

# Create the choropleth map
plot_ly(
  data = diabetes_by_state,
  type = "choropleth",
  locations = ~abb,  # Use two-letter state/territory abbreviations
  z = ~prevalence,  # Color intensity based on prevalence
  locationmode = "USA-states",  # Map mode for U.S. states and territories
  colorscale = "Viridis",  # Use the Viridis color scale
  colorbar = list(title = "Prevalence (%)"),
  zmin = 0,  # Set minimum value for the color scale
  zmax = max(diabetes_by_state$prevalence, na.rm = TRUE),  # Set max value
  showscale = TRUE  # Show the color scale
) |> 
  layout(
    title = "Diabetes Prevalence Across the U.S. and Territories",
    geo = list(
      scope = "usa",  # Focus on U.S. states and territories
      projection = list(type = "albers usa"),  # Standard USA projection
      showlakes = TRUE,  # Show lakes
      lakecolor = "lightblue"  # Set lake color
    )
  ) 
```

Above is a map of the distribution of the prevalence of diabetes in the US, where the lowest prevalence is in Colorado (9.3%), and the highest prevalence of diabetes is in West Virginia (20.42%), assuming that the survey responses are representative of the general US population. There are no available data for Kentucky and Pennsylvania. In general, the South has the highest overall prevalence of diabetes in the U.S.

# Sex at Birth and Diabetes

Overall, diabetes diagnosis (as well as type) is pretty evenly spread between males and females (at birth).

#### `has_diabetes` comparison

```{r has_compare, warning=FALSE, message=FALSE}
diabetes_df |> 
  group_by(has_diabetes, sex_at_birth) |> 
  summarize(has_diabetes_count = n()) |> 
  pivot_wider(
    names_from = has_diabetes,
    values_from = has_diabetes_count
  ) |> 
  rename(`Sex at Birth` = sex_at_birth) |> 
  knitr::kable()
```

#### `diab_type` comparison

```{r diab_compare, warning=FALSE, message=FALSE}
diabetes_df |> 
  group_by(diab_type, sex_at_birth) |> 
  summarize(diab_type_count = n()) |> 
  pivot_wider(
    names_from = diab_type,
    values_from = diab_type_count
  ) |> 
  rename(`Sex at Birth` = sex_at_birth) |> 
  knitr::kable()
```

#### `eval_type` comparison

```{r eval_compare, warning=FALSE, message=FALSE}
diabetes_df |> 
  group_by(eval_type, sex_at_birth) |> 
  summarize(eval_type_count = n()) |> 
  pivot_wider(
    names_from = eval_type,
    values_from = eval_type_count
  ) |> 
  rename(`Sex at Birth` = sex_at_birth) |> 
  knitr::kable()
```

# Age and Diabetes Diagnosis

In the graph below, we see that as age increases, the count of diabetes cases increase (using the `has_diabetes` variable), with the highest number of cases being in the 65-74 age group. The number of diabetes cases decrease slightly after age 74. 

```{r age_diab, warning=FALSE, message=FALSE}
diabetes_df |> 
  mutate(
    age_category = factor(age_category, levels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"), order = TRUE)
  ) |> 
  group_by(age_category) |> 
  summarize(count = sum(has_diabetes == "Diabetic", na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = age_category, y = count, fill = age_category)) +
  geom_bar(stat = "identity", color = "black") +
  labs(
    title = "Distribution of Diabetes among Age Groups Surveyed",
    x = "Age Category (5 years)",
    y = "Count of Reported Diabetes Cases",
    fill = "Age Category"
  ) +
  theme_minimal()
```

# Age of Onset and Type of Diabetes


Comparing the two measures for type of diabetes (`eval_type` and `diab_type`), we don't see as many type I diagnoses for `eval_type`, which makes sense because of how we evaluate the variable. Since we categorize everyone over the age of 30 as having T2D, we must **proceed in our analysis with caution**, knowing that we are actively removing a small portion of type 1 cases for each year above 30.

*Due to time constraints, we did not impute, but based on the results here, it may be useful to randomly impute a few type 1 cases per age group using these counts/percentages.*

```{r report_diab, message=FALSE, warning=FALSE}
diabetes_df |> 
  filter(!(is.na(diab_type))) |> 
  group_by(age_onset, diab_type) |> 
  summarize(
    count = n(),  # Count the number of rows for each combination
    .groups = "drop"
  ) |> 
  ggplot(aes(x = age_onset, y = count, fill = diab_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Reported Diabetes Type by Age of Onset",
    x = "Age of Onset",
    y = "Count",
    fill = "Diabetes Type"
  ) +
  theme_minimal() +
  theme(legend.position="bottom")
```

```{r onset_eval, message=FALSE, warning=FALSE}
diabetes_df |> 
  group_by(age_onset, eval_type) |> 
  summarize(
    count = n(),  # Count the number of rows for each combination
    .groups = "drop"
  ) |> 
  filter(!(eval_type == "Not diabetic")) |> 
  ggplot(aes(x = age_onset, y = count, fill = eval_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Evaluated Diabetes Type by Age of Onset",
    x = "Age of Onset",
    y = "Count",
    fill = "Diabetes Type"
  ) +
  theme_minimal() +
  theme(legend.position="bottom")
```

