---
title: "Data Cleaning"
author: "mk4995"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

Import the data.

```{r}
diabetes_data = read_csv("data/diabetes_data.csv")
```

The initial data contains 433,323 rows and 34 columns.

We want to start by recoding the values that are not clear answers, "don't knows", refusals, and missing ALL as missing data. We will also recode numeric variables into factor variables.

```{r}
# Define a helper function for recoding
recode_binary = function(var) {
  case_when(
    var == 1 ~ 1,
    var == 2 ~ 0,
    var %in% c(7, 9, NA) ~ NA
  )
}
```


```{r}
diabetes_recoded_data = 
  diabetes_data |> 
  mutate(
    has_diabetes = case_when( # yes = 2, no = 0, pre-diabetes = 1, gestational = 3
      diabete4 == 1 ~ 2,
      diabete4 == 3 ~ 0,
      diabete4 == 4 ~ 1,
      diabete4 == 2 ~ 3,
      diabete4 %in% c(7, 9, NA) ~ NA
    ),
    age_onset = case_when(
      diabage4 >= 1 & diabage4 <= 97 ~ diabage4, 
      diabage4 %in% c(98, 99, NA) ~ NA
    ),
    diab_type = case_when(
      diabtype == 1 ~ diabtype,
      diabtype == 2 ~ diabtype, 
      diabtype %in% c(7, 9, NA) ~ NA
    ),
    prediabetic = case_when( # yes = 1, no = 0, gestational = 2
      prediab2 == 1 ~ 1,
      prediab2 == 2 ~ 2,
      prediab2 == 3 ~ 1,
      prediab2 %in% c(7, 9, NA) ~ NA
    ),
    insulin = recode_binary(insulin1),
    high_chol = recode_binary(toldhi3),
    covid_test = recode_binary(covidpo1),
    high_bp = case_when( # yes = 2, borderline = 1, no = 0, gestational = 3
      bphigh6 == 1 ~ 2,
      bphigh6 == 4 ~ 1,
      bphigh6 == 3 ~ 0,
      bphigh6 == 2 ~ 3, 
      bphigh6 %in% c(7, 9, NA) ~ NA
    ),
    high_bs = case_when(
      pdiabts1 %in% c(1, 2, 3, 4, 5, 6) ~ 1,
      pdiabts1 == 8 ~ 0,
      pdiabts1 %in% c(7, 9, NA) ~ NA
    ),
    a1c_check = case_when(
      chkhemo3 <= 76 ~ 1,
      chkhemo3 == 88 ~ 0,
      chkhemo3 %in% c(77, 98, 99, NA) ~ NA
    ),
    any_exercise = case_when(
      exerany2 == 1 ~ 1,
      exerany2 == 2 ~ 0,
      exerany2 == 9 ~ NA
    ),
    heart_attack = recode_binary(cvdinfr4),
    chd = recode_binary(cvdcrhd4),
    stroke = recode_binary(cvdstrk3),
    asthma_ever = recode_binary(asthma3),
    asthma_now = recode_binary(asthnow),
    bronchitis = recode_binary(chccopd3),
    depression = recode_binary(addepev3),
    kidney_disease = recode_binary(chckdny2),
    pregnant = recode_binary(pregnant),
    arthritis = recode_binary(havarth4),
    smoker = recode_binary(smoke100),
    state = cdlTools::fips(state, to = "Name"),
    educa = ifelse(educa == 9, NA, educa),
    education = factor(educa,
      levels = c(1, 2, 3, 4, 5, 6),
      labels = c(
        "Kindergarten or less",
        "Elementary",
        "Some high school",
        "High school graduate",
        "Some college or technical school",
        "College graduate"
      )
    ),
    sex_at_birth = ifelse(sexvar == 1, 1,       # Male = 1
                        ifelse(sexvar == 2, 0, NA)),
    weight_lbs = case_when(
      weight2 >= 50 & weight2 <= 776 ~ weight2, # Pounds are already in the range
      weight2 >= 9023 & weight2 <= 9352 ~ (weight2 - 9000) * 2.20462, # Convert kg to lbs
      weight2 %in% c(7777, 9999, NA) ~ NA, # Don’t know/Not sure
    ),
    height = case_when(
      height3 >= 200 & height3 <= 711 ~ height3,
      height3 >= 9061 & height3 <= 9998 ~ {
        cm_height <- height3 - 9000                       # Extract cm
        total_inches <- round(cm_height * 0.393701)       # Convert to inches
        feet <- floor(total_inches / 12)                 # Get feet
        inches <- total_inches %% 12                     # Get remaining inches
        as.numeric(sprintf("%d%02d", feet, inches))     
      },
      height3 %in% c(7777, 9999, NA) ~ NA,
    ),
    urbrural = case_when(
      urbstat == 1 ~ 1,
      urbstat == 2 ~ 0,
      urbstat == NA ~ NA
    ),
    smoker_now = case_when( # now = 1, former = 2, never = 0
      smoker3 %in% c(1, 2) ~ 1,
      smoker3 == 3 ~ 2,
      smoker3 == 4 ~ 0,
      smoker3 == 9 ~ NA
    ),
    ageg5yr = ifelse(ageg5yr == 14, NA, ageg5yr),
    age_category = factor(
      ageg5yr,
      levels = c(1:13),  # Define the levels in the natural order
      labels = c(
        "18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64",
        "65-69", "70-74", "75-79", "80+"),
      ordered = TRUE  # Specify that this is an ordered factor
    )
  ) |> 
  select(has_diabetes, age_onset, diab_type, prediabetic, insulin, high_chol, covid_test, high_bp, high_bs, a1c_check, any_exercise, heart_attack, chd, stroke, asthma_ever, asthma_now, bronchitis, depression, kidney_disease, pregnant, arthritis, smoker, state, education, sex_at_birth, weight_lbs, height, urbrural, smoker_now, age_category)
  
```

The final dataset has the following variables.

```{r}
colnames(diabetes_recoded_data)
```

We can also check that our code ran as expected by checking if there are any columns that contain only NA values.

```{r}
all_na_columns <- colnames(diabetes_recoded_data)[colSums(is.na(diabetes_recoded_data)) == nrow(diabetes_recoded_data)]

# Print the names of columns with all NA values
if (length(all_na_columns) > 0) {
  cat("Columns with all NA values:\n")
  print(all_na_columns)
} else {
  cat("No columns with all NA values found.\n")
}
```

Export the cleaned dataset (in case it's needed in other pages).

```{r}
write_csv(diabetes_recoded_data, "data/cleaned_diabetes_data.csv")
```

