---
title: "Exploring Related Factors and Comorbidities of T2D"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(rlang)
library(viridis)

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


```{r load_data, message = FALSE}
# Load and wrangle the overall data.
diabetes_df = read_csv("data/cleaned_diabetes_data.csv")

# Binary helper function
recode_factor = function(var) {
  factor(var, levels = c("No", "Yes"))
}

# Clean the diabetes variables
diabetes_df = 
  diabetes_df |> 
  mutate(
    has_diabetes = factor(has_diabetes, levels = c("Not diabetic", "Pre-diabetic", "Diabetic", "Gestational")),
    sex_at_birth = fct_infreq(sex_at_birth),
    age_category = factor(age_category, levels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"), order = TRUE),
    pregnant = recode_factor(pregnant),
    eval_type = factor(eval_type, levels = c("Not diabetic", "Type 1", "Type 2")),
    diab_type = factor(diab_type, levels = c("Type 1", "Type 2"))
  )
```

# Demographics {.tabset}
```{r demographics_df}
diabetes_demographics = 
  diabetes_df |> 
  mutate(
    good_health = recode_factor(good_health),
    physical_activity = factor(
      physical_activity,
      levels = c(
        "Highly active",
        "Active",
        "Insufficiently active",
        "Inactive"
      )
    ),
    urb_rural = factor(urb_rural, levels = c("Urban", "Rural")),
    race = as.factor(race),
    income = factor(
      income,
      levels = c(
        "Less than 15K",
        "15K=25K",
        "25-35K",
        "35K-50K",
        "50K-100K",
        "100K-200K",
        "More than 200K"
      )
    ),
    education = factor(education,
      levels = c(
        "Kindergarten or less",
        "Elementary",
        "Some high school",
        "High school graduate",
        "Some college or technical school",
        "College graduate"
      )
    )
  ) |> 
  select(has_diabetes, diab_type, eval_type, good_health, physical_activity, urb_rural, race, income, education) |> 
  rename(
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type,
    `In Good Health` = good_health,
    `Level of Physical Activity` = physical_activity,
    `Urban/Rural Status` = urb_rural,
    `Race/Ethnicity` = race,
    `Income Level` = income,
    `Educational Attainment` = education
  )
```

This section explores the distribution of demographic information among diabetes diagnoses.

- Whether the respondent is in good health
- Level of physical activity
- Living in an urban vs. rural area
- Race/Ethnicity
- Income level
- Educational attainment

```{r diabetes_plots}
# Define a function for the plot

comorbidities_plot = function(df, comorbidity, diabetes) {
  # Convert character strings to symbols
  comorbidity_sym <- rlang::sym(comorbidity)
  diabetes_sym <- rlang::sym(diabetes)
  
  df |> 
    filter(!is.na(!!comorbidity_sym), !(is.na(!!diabetes_sym))) |> 
    group_by(!!comorbidity_sym, !!diabetes_sym) |> 
    summarize(n = n(), .groups = 'drop') %>%
    group_by(!!diabetes_sym) %>%
    mutate(Percent = n / sum(n) * 100) |> 
    ggplot(aes(x = !!diabetes_sym, y = Percent, fill = !!comorbidity_sym)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", Percent)),
              position = position_dodge(width = 0.9),
              size = 2.5, 
              vjust = -0.5) + 
    labs(
      title = str_c(diabetes, " by ", comorbidity), 
      x = diabetes, 
      y = "Percent (%)", 
      fill = comorbidity) +
    scale_fill_viridis_d(na.value = "gray50", option = "D") +
    theme_minimal() +
    theme(legend.position="bottom")
}
```

## Good Health {.tabset}

```{r}
demo = "In Good Health"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Physical Activity {.tabset}

```{r}
demo = "Level of Physical Activity"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Urban/Rural {.tabset}

```{r}
demo = "Urban/Rural Status"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Race/Ethnicity {.tabset}

```{r}
demo = "Race/Ethnicity"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Income {.tabset}

```{r}
demo = "Income Level"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Education {.tabset}

```{r}
demo = "Educational Attainment"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

# Comorbidities {.tabset}

```{r comorbidities_df}
diabetes_comorbidities = 
  diabetes_df |> 
  mutate(
    kidney_disease = recode_factor(kidney_disease),
    depression = recode_factor(depression),
    arthritis = recode_factor(arthritis),
    michd = recode_factor(michd),
    blind = recode_factor(blind),
    covid = recode_factor(covid),
    stroke = recode_factor(stroke),
    bronchitis = recode_factor(bronchitis),
    asthma_ever = recode_factor(asthma_ever),
    asthma_now = recode_factor(asthma_now)
  ) |> 
  select(has_diabetes, diab_type, eval_type, kidney_disease, arthritis, depression, michd, blind, covid, stroke, bronchitis, asthma_ever, asthma_now) |> 
  rename(
    `History of Kidney Disease` = kidney_disease,
    `History of MI/CHD` = michd,
    `History of Depression` = depression,
    `Blindness` = blind,
    `History of Stroke` = stroke,
    `History of Arthritis` = arthritis,
    `History of Asthma` = asthma_ever,
    `Currently Have Asthma` = asthma_now,
    `Ever Tested Positive for Covid-19` = covid,
    `History of Bronchitis` = bronchitis,
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type
  )
```

This section explores the distribution of co-morbidities across diabetes diagnoses.

- History of kidney disease
- History of depression
- Blindness
- History of myocardial infarction (MI) or coronary heart disease (CHD)
- History of stroke
- History of arthritis
- History of asthma
- Currently has asthma
- Ever tested positive for COVID-19
- History of bronchitis

## Kidney disease {.tabset}

```{r}
comorbidity = "History of Kidney Disease"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Depression {.tabset}

```{r}
comorbidity = "History of Depression"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Blindness {.tabset}

```{r}
comorbidity = "Blindness"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## MI/CHD {.tabset}

```{r}
comorbidity = "History of MI/CHD"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Stroke {.tabset}

```{r}
comorbidity = "History of Stroke"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Arthritis {.tabset}

```{r}
comorbidity = "History of Arthritis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Asthma {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Has Diabetes")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Reported Type")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Evaluated Type")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Evaluated Type")
```

## COVID-19 {.tabset}

```{r}
comorbidity = "Ever Tested Positive for Covid-19"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Bronchitis {.tabset}

```{r}
comorbidity = "History of Bronchitis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```


# Associated Factors {.tabset}

```{r diabetes_factors}
diabetes_factors = 
  diabetes_df |> 
  mutate(
    high_bp = recode_factor(high_bp),
    a1c_check = case_match(a1c_check, 0 ~ "No", 1 ~ "Yes", NA ~ NA),
    a1c_check = recode_factor(a1c_check),
    chol_meds = recode_factor(chol_meds),
    chol_check = recode_factor(chol_check),
    smoker = factor(smoker, levels = c("Current smoker (every day)", "Current smoker (some days)", "Former smoker", "Never smoked")),
    ecigs = recode_factor(ecigs),
    prediabetic = factor(prediabetic, levels = c("Yes", "No", "Gestational")),
    obese = recode_factor(obese),
    cancer = recode_factor(cancer),
    heavy_drinking = recode_factor(heavy_drinking),
    binge_drinking = recode_factor(binge_drinking)
  ) |> 
  select(high_bp, a1c_check, chol_meds, chol_check, smoker, ecigs, prediabetic, obese, cancer, heavy_drinking, binge_drinking, has_diabetes, diab_type, eval_type) |> 
  rename(
    `High Blood Pressure` = high_bp,
    `High Cholesterol` = chol_check,
    `Meds Usage for Cholesterol` = chol_meds,
    `Checked A1C in the Past Year` = a1c_check,
    `Smoking Status` = smoker,
    `E-Cigarette Use` = ecigs,
    `Prediabetes Diagnosis` = prediabetic,
    `Overweight/Obese (Computed)` = obese,
    `Cancer Diagnosis` = cancer,
    `Heavy Drinking` = heavy_drinking,
    `Binge Drinking` = binge_drinking,
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type
  )
```

This section explores the distribution of related factors across diabetes diagnoses.

- High blood pressure
- High blood sugar
- High cholesterol
- Checked for A1C over the past year
- Smoking status
- Ever received a prediabetes diagnosis
- Overweight/obese
- Ever received a cancer diagnosis
- Drinking behavior

## High BP {.tabset}

```{r}
factor = "High Blood Pressure"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## High Cholesterol {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Has Diabetes")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Reported Type")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Evaluated Type")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Evaluated Type")
```

## Smoker {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Has Diabetes")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Reported Type")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Evaluated Type")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Evaluated Type")
```

## Obesity {.tabset}

```{r}
factor = "Overweight/Obese (Computed)"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## Cancer {.tabset}

```{r}
factor = "Cancer Diagnosis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## Drinking {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Has Diabetes")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Reported Type")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Evaluated Type")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Evaluated Type")
```

