---
title: "Exploring Related Factors and Comorbidities of T2D"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(rlang)
library(viridis)

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

In this section, we explored the relationship between lifestyle, health factors, and comorbidities and the different possible diagnoses of diabetes in our data. We compared each factor to one of our three outcome variables, `has_diabetes`, `diab_type`, and `eval_type`, to identify the overarching relationship between specific demographics and diabetes, along with similarities and differences between the various kinds of diagnosis.

```{r load_data, message = FALSE}
# Load and wrangle the overall data.
diabetes_df = read_csv("data/cleaned_diabetes_data.csv")

# Binary helper function
recode_factor = function(var) {
  factor(var, levels = c("No", "Yes"))
}

# Clean the diabetes variables
diabetes_df = 
  diabetes_df |> 
  mutate(
    has_diabetes = factor(has_diabetes, levels = c("Not diabetic", "Pre-diabetic", "Diabetic", "Gestational")),
    sex_at_birth = fct_infreq(sex_at_birth),
    age_category = factor(age_category, levels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"), order = TRUE),
    pregnant = recode_factor(pregnant),
    eval_type = factor(eval_type, levels = c("Not diabetic", "Type 1", "Type 2")),
    diab_type = factor(diab_type, levels = c("Type 1", "Type 2"))
  )
```

# Demographics {.tabset}
```{r demographics_df}
diabetes_demographics = 
  diabetes_df |> 
  mutate(
    good_health = recode_factor(good_health),
    physical_activity = factor(
      physical_activity,
      levels = c(
        "Highly active",
        "Active",
        "Insufficiently active",
        "Inactive"
      )
    ),
    urb_rural = factor(urb_rural, levels = c("Urban", "Rural")),
    race = as.factor(race),
    income = factor(
      income,
      levels = c(
        "Less than 15K",
        "15K=25K",
        "25-35K",
        "35K-50K",
        "50K-100K",
        "100K-200K",
        "More than 200K"
      )
    ),
    education = factor(education,
      levels = c(
        "Kindergarten or less",
        "Elementary",
        "Some high school",
        "High school graduate",
        "Some college or technical school",
        "College graduate"
      )
    )
  ) |> 
  select(has_diabetes, diab_type, eval_type, good_health, physical_activity, urb_rural, race, income, education) |> 
  rename(
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type,
    `In Good Health` = good_health,
    `Level of Physical Activity` = physical_activity,
    `Urban/Rural Status` = urb_rural,
    `Race/Ethnicity` = race,
    `Income Level` = income,
    `Educational Attainment` = education
  )
```

This section explores the distribution of demographic information among diabetes diagnoses.

- Whether the respondent is in good health
- Level of physical activity
- Living in an urban vs. rural area
- Race/Ethnicity
- Income level
- Educational attainment

```{r diabetes_plots}
# Define a function for the plot

comorbidities_plot = function(df, comorbidity, diabetes) {
  # Convert character strings to symbols
  comorbidity_sym <- rlang::sym(comorbidity)
  diabetes_sym <- rlang::sym(diabetes)
  
  df |> 
    filter(!is.na(!!comorbidity_sym), !(is.na(!!diabetes_sym))) |> 
    group_by(!!comorbidity_sym, !!diabetes_sym) |> 
    summarize(n = n(), .groups = 'drop') %>%
    group_by(!!diabetes_sym) %>%
    mutate(Percent = n / sum(n) * 100) |> 
    ggplot(aes(x = !!diabetes_sym, y = Percent, fill = !!comorbidity_sym)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", Percent)),
              position = position_dodge(width = 0.9),
              size = 2.5, 
              vjust = -0.5) + 
    labs(
      title = str_c(diabetes, " by ", comorbidity), 
      x = diabetes, 
      y = "Percent (%)", 
      fill = comorbidity) +
    scale_fill_viridis_d(na.value = "gray50", option = "D") +
    theme_minimal() +
    theme(legend.position="bottom")
}
```

People with better health and physical activity practices have fewer diabetes diagnoses. Between type I and type II diagnoses, physical activity is about the same, based on the reported type. On the other hand, type II diagnoses have a higher percentage of inactive people, based on the evaluated type.

Urban/rural status is pretty similar acoss the board, but that may be because the overall percentage of respondents is unevenly distributed between urban and rural populations. We would suggest checking that distribution. 

We see a similar case for race and ethnicity - a weighting procedure may better show if there is truly a difference in diabetes diagnosis between different races/ethnicities. We see that a higher percentage of Hispanic respondents had type I diabetes compared to type II in both the reported and evaluated type comparisons, whereas white respondents were more likely to have type II diabetes than type I diabetes.

While we don’t see any major differences jumps in diabetes diagnosis, there are a few things to note:

People making 50K-100K make up a large portion of the survey respondents
According to the reported type, people making 50-100K made up a larger portion of the type II respondents than they did type I
However, that looks relatively the same in the evaluated type comparison

People with (at least) a college degree make up a larger proportion of the sample of people who are not diabetic than the sample of people who are diabetic. Surprisingly, accordiing to the evaluated type variable, a significantly larger percentage of people who graduated high school have type I diabetes, compared to the percentage of people who graduated high school with type II diabetes.

## Good Health {.tabset}

```{r}
demo = "In Good Health"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Physical Activity {.tabset}

```{r}
demo = "Level of Physical Activity"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Urban/Rural {.tabset}

```{r}
demo = "Urban/Rural Status"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Race/Ethnicity {.tabset}

```{r}
demo = "Race/Ethnicity"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Income {.tabset}

```{r}
demo = "Income Level"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

## Education {.tabset}

```{r}
demo = "Educational Attainment"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_demographics, demo, "Evaluated Type")
```

# Comorbidities {.tabset}

```{r comorbidities_df}
diabetes_comorbidities = 
  diabetes_df |> 
  mutate(
    kidney_disease = recode_factor(kidney_disease),
    depression = recode_factor(depression),
    arthritis = recode_factor(arthritis),
    michd = recode_factor(michd),
    blind = recode_factor(blind),
    covid = recode_factor(covid),
    stroke = recode_factor(stroke),
    bronchitis = recode_factor(bronchitis),
    asthma_ever = recode_factor(asthma_ever),
    asthma_now = recode_factor(asthma_now)
  ) |> 
  select(has_diabetes, diab_type, eval_type, kidney_disease, arthritis, depression, michd, blind, covid, stroke, bronchitis, asthma_ever, asthma_now) |> 
  rename(
    `History of Kidney Disease` = kidney_disease,
    `History of MI/CHD` = michd,
    `History of Depression` = depression,
    `Blindness` = blind,
    `History of Stroke` = stroke,
    `History of Arthritis` = arthritis,
    `History of Asthma` = asthma_ever,
    `Currently Have Asthma` = asthma_now,
    `Ever Tested Positive for Covid-19` = covid,
    `History of Bronchitis` = bronchitis,
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type
  )
```

This section explores the distribution of co-morbidities across diabetes diagnoses.

- History of kidney disease
- History of depression
- Blindness
- History of myocardial infarction (MI) or coronary heart disease (CHD)
- History of stroke
- History of arthritis
- History of asthma
- Currently has asthma (only asked of people who answered yes to history of asthma)
- Ever tested positive for COVID-19
- History of bronchitis

A larger proportion of people with diabetes also have kidney disease than other groups. The reported type shows no significant difference in diagnosis across type of diabetes, but the evaluated type shows that people with kidney disease are more highly represented in the group with type II diabetes.

Diabetes has no significant difference between groups who self-reported diabetes status, but according to our evaluated type, people with type I diabetes reported having depression more than other groups.

Blindness is slightly more common in people who have diabetes compared to people who are not diabetic, but we see some opposition between reported and evaluated type I and II diabetes status.

History of MI/CHD is more common in diabetics than non-diabetics. Reported type shows no difference, but evaluated status shows a significant increase in the percentage of people with MI/CHD in the type II group. Stroke is similar but less significant.

Arthritis is more common in diabetics than non-diabetics. There seems to be a significant difference in type I vs. type II in both reported and evaluated diagnoses, but the evaluated type has a much more significant difference between groups.

Since the people who currently have asthma are also included in the group of people who have a history of asthma, we will focus on the latter. A history of asthma does not seem to have a significant relationship across diabetes diagnoses. 

Getting COVID-19 is not super related to diagnosis, but there is a slightly lower percentage of positive tests in the diabetic group compared to non-diabetic. The evaluated type has a significantly higher percentage of positive covid-19 diagnoses in type I than type II.

Bronchitis does not have any significant relationships except for the evaluated type, where there is a higher percentage of people with bronchitis in the type II group than type I.

## Kidney disease {.tabset}

```{r}
comorbidity = "History of Kidney Disease"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Depression {.tabset}

```{r}
comorbidity = "History of Depression"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Blindness {.tabset}

```{r}
comorbidity = "Blindness"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## MI/CHD {.tabset}

```{r}
comorbidity = "History of MI/CHD"
```

### `has_diabetes`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Stroke {.tabset}

```{r}
comorbidity = "History of Stroke"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Arthritis {.tabset}

```{r}
comorbidity = "History of Arthritis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Asthma {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Has Diabetes")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Reported Type")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, "History of Asthma", "Evaluated Type")
comorbidities_plot(diabetes_comorbidities, "Currently Have Asthma", "Evaluated Type")
```

## COVID-19 {.tabset}

```{r}
comorbidity = "Ever Tested Positive for Covid-19"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

## Bronchitis {.tabset}

```{r}
comorbidity = "History of Bronchitis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_comorbidities, comorbidity, "Evaluated Type")
```

# Associated Factors {.tabset}

```{r diabetes_factors}
diabetes_factors = 
  diabetes_df |> 
  mutate(
    high_bp = recode_factor(high_bp),
    a1c_check = case_match(a1c_check, 0 ~ "No", 1 ~ "Yes", NA ~ NA),
    a1c_check = recode_factor(a1c_check),
    chol_meds = recode_factor(chol_meds),
    chol_check = recode_factor(chol_check),
    smoker = factor(smoker, levels = c("Current smoker (every day)", "Current smoker (some days)", "Former smoker", "Never smoked")),
    ecigs = recode_factor(ecigs),
    prediabetic = factor(prediabetic, levels = c("Yes", "No", "Gestational")),
    obese = recode_factor(obese),
    cancer = recode_factor(cancer),
    heavy_drinking = recode_factor(heavy_drinking),
    binge_drinking = recode_factor(binge_drinking)
  ) |> 
  select(high_bp, a1c_check, chol_meds, chol_check, smoker, ecigs, prediabetic, obese, cancer, heavy_drinking, binge_drinking, has_diabetes, diab_type, eval_type) |> 
  rename(
    `High Blood Pressure` = high_bp,
    `High Cholesterol` = chol_check,
    `Meds Usage for Cholesterol` = chol_meds,
    `Checked A1C in the Past Year` = a1c_check,
    `Smoking Status` = smoker,
    `E-Cigarette Use` = ecigs,
    `Prediabetes Diagnosis` = prediabetic,
    `Overweight/Obese (Computed)` = obese,
    `Cancer Diagnosis` = cancer,
    `Heavy Drinking` = heavy_drinking,
    `Binge Drinking` = binge_drinking,
    `Has Diabetes` = has_diabetes,
    `Reported Type` = diab_type,
    `Evaluated Type` = eval_type
  )
```

This section explores the distribution of related factors across diabetes diagnoses.

- High blood pressure
- High cholesterol
- Smoking status
- Ever received a prediabetes diagnosis
- Overweight/obese
- Ever received a cancer diagnosis
- Drinking behavior

We originally explored high blood sugar and A1C checks in this section, but decided not to publish those models because only people with diabetes (as per `has_diabetes`) were asked this question, severely skewing our results.

People with low/normal blood pressure were represented much more in the group of people who are diabetic compared to any other group. There is a slight increase in high blood representation percentage in the type I group compared to the type II group in the reported cases, but that is heightened in comparing the evaluated type variable.

High cholesterol is more common in the non-diabetic group than the diabetic group. There is a higher percentage of high cholesterol in type I patients than type II.

In our smoking comparison, we examined both tobacco and e-cigarette users. Former smokers make up a higher percentage of the diabetic group than the non-diabetic group, with a higher percentage in the type II group.

People who have been computed to be overweight/obese by BRFSS investigators are more represented in the type I group than the type II group. Cancer is significantly more associated with type II than type I diagnosis, according to the evaluated type variable. Drinking does not have a significant association with any of the data.

## High BP {.tabset}

```{r}
factor = "High Blood Pressure"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## High Cholesterol {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Has Diabetes")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Reported Type")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "High Cholesterol", "Evaluated Type")
comorbidities_plot(diabetes_factors, "Meds Usage for Cholesterol", "Evaluated Type")
```

## Smoker {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Has Diabetes")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Reported Type")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "Smoking Status", "Evaluated Type")
comorbidities_plot(diabetes_factors, "E-Cigarette Use", "Evaluated Type")
```

## Obesity {.tabset}

```{r}
factor = "Overweight/Obese (Computed)"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## Cancer {.tabset}

```{r}
factor = "Cancer Diagnosis"
```

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, factor, "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, factor, "Evaluated Type")
```

## Drinking {.tabset}

### `has_diabetes` {.tabset}

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Has Diabetes")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Has Diabetes")
```

### `diab_type`

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Reported Type")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Reported Type")
```

### `eval_type`

```{r}
comorbidities_plot(diabetes_factors, "Heavy Drinking", "Evaluated Type")
comorbidities_plot(diabetes_factors, "Binge Drinking", "Evaluated Type")
```
