---
title: "Regression"
author: "Arghya Kannadaguli (ak5357)"
date: "2024-12-03"
output: html_document
---

```{r include = FALSE}
# Load libraries
library(tidyverse)
library(purrr)
library(glmnet)
library(dplyr)
library(modelr)

# Do not eval anything, when the file is knit
knitr::opts_chunk$set(eval = FALSE)
```

We need to figure out a more efficient way of getting the cleaned data (with in-tact data types) into this Rmd. Perhaps we can transfer the data cleaning code to a .R script instead of .Rmd, and use the `source()` function.

Also **NOTE TO MAYA**

* We need to encode `smoker` variable in the `data_cleaning.Rmd`. I've done it here for now.
* diab_type should be a factor variable

```{r}
rmarkdown::render("data_cleaning.Rmd")
diabetes_df = diabetes_recoded |> 
  mutate(
    smoker = fct_relevel(smoker, c("Never smoked", "Former smoker", 
                                   "Current smoker (some days)", "Current smoker (every day)")),
    diab_type = as.factor(diab_type))
```

Get all non-outcome variables
```{r}
all_vars = 
  diabetes_df |> 
  select(-c(diab_type, has_diabetes, eval_type, prediabetic, where(is.character))) |> 
  colnames()

all_vars_str = all_vars |> 
  paste(collapse = " + ")

formula = reformulate(var, response = "diab_type")
```

TESTING to see which variables we can plug in a glm model, and which ones throw errors due to presence of NA values. This for loop has issues btw.
```{r eval = FALSE}

for (i in 1:length(all_vars)){
  var = all_vars[[i]]
  print(paste("Trying variable:", var))
  formula = reformulate(var, response = "diab_type")
  
  diabetes_df |> 
    filter(!is.na(!!var)) |> 
    glm(formula, data = _, family = binomial()) |> 
    broom::tidy()
}
```

Exports a table with the number of null values in each column. Results looking bleak.
```{r eval = FALSE}
diabetes_df |> 
  summarize(
    across(everything(), ~ sum(is.na(.)), .names = "{col}")) |> 
      pivot_longer(
        everything(),
        names_to = "colname",
        values_to = "null_count"
      ) |> 
  mutate(null_prop = null_count / nrow(diabetes_df)) |> 
  arrange(null_prop) |> 
  write_csv("variable_null_counts.csv")
```

